\documentclass{report}
\usepackage{suthesis-2e}
\usepackage{hyperref}
\tablespagefalse%


\dept{Computer Science}

\include{macros}

\begin{document}
\title{Backpack: Towards Practical Mix-in Linking in Haskell \\ REVISION 2017--05--22}
\author{Edward Z. Yang}
\principaladviser{David Mazi\`eres}
\firstreader{Derek Dreyer}
\secondreader{John Mitchell}
%\thirdreader{Jane Supernumerary} %if needed
%\fourthreader{Severus Snape} %if needed

\beforepreface%
%\prefacesection{Preface}
%    (To be written)
\prefacesection{Acknowledgments}
    I want to thank David Mazi\`eres for his help, advice, and
    willingness to let me go off and explore the strange world
    of module systems for Haskell.  He always stepped up to the bat
    when I needed it.  I also have to thank my coadvisor John
    Mitchell for his sage advice and his CS242 syllabus---without
    which I would have lost much more sleep teaching at Stanford.

    This thesis would not exist without the work of Scott Kilpatrick.
    Backpack is really two PhDs of work: Scott did the theory, I did the
    implementation.  He really is my academic better half.  I also have
    to thank Derek Dreyer who acted as a remote technical advisor
    while I was in the weeds.
    I would not have come to do a PhD in the first place without the
    superb mentorship of Simon Peyton Jones and Simon Marlow.  They are,
    and continue to be, an inspiration to me.

    Many individuals both inside and outside of Stanford have helped
    make the PhD journey a little less lonely.  At risk of forgetting
    someone, they include Deian Stefan, Ali Mashtizadeh, Sergio Benitez,
    Amit Levy, David Terei, Adam Belay, Daniel Giffin, Alejandro Russo,
    Petr Marchenko, Henry Corrigan-Gibbs, Riad Wahby, Danae Metaxca,
    Stefan Heule, Ryan Newton, Giovanni Campagna, Andreas Rossberg and
    Mary-Jane Swenson.

    I would also like to thank Mom, Dad, and my two brothers, James
    and David, for being a rock of unconditional support and love,
    both before and during my PhD.

    And last but not least, I would like to thank Kaixi Ruan, love
    of my life.  This one's for you.
\afterpreface%

\input{intro}
\input{tour}
\input{overview}
\input{uid}
\input{mixin}
\input{compiler}
\input{implementation}
\input{evaluation}
\input{limitations}
\input{related}
\input{conclusion}
\input{appendix}

\bibliographystyle{plain}
\bibliography{references}
\end{document}
